---
title: "Conditional Prediction of L3 Offence Subtypes given L1 + L2"
subtitle: "With Spatial Feature Engineering — k-means Crime Zones"
author: "CompStat Project 2 — W06G03"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 6, dpi = 110,
  cache = TRUE,
  cache.lazy = FALSE
)
options(scipen = 999)
```

# Section 0 — Plan & risk register {#plan}

The research question is:

> *To what extent do temporal (calendar/holiday), spatial (suburb), and
> socio-economic (SEIFA decile) features improve conditional prediction of L3
> offence subtypes, **given** L1 and L2?*

**Key difference from the base version:** `Suburb_Incident` is never used
directly as a model feature. Instead, all **965 raw suburbs** are individually
clustered into a 10-level factor `Zone_km` using k-means on suburb-level
density and SEIFA features. Every suburb — even one with 3 rows — gets its
own zone assignment based on its actual SEIFA deciles, not a generic
`"OTHER"` bucket. This eliminates the unseen-level NA risk, reduces the
Spatial multinomial from ~9,828 weights to 280, and cuts fit time from
**~35 min to ~2 min** per block.

## Data quirks that motivate defensive coding

* Column names contain spaces/hyphens — renamed to `Suburb_Incident`, `L1`,
  `L2`, `L3` immediately on load.
* All 965 raw suburbs are retained — **no `"OTHER"` collapsing** in this
  file because `Suburb_Incident` is only a grouping key for zone derivation,
  never a direct model predictor.
* SEIFA missing for ~873 rows → imputed by suburb median (on the original
  965 suburb groups), then column-wide median fallback.
* Duplicate rows are **kept** — they are real repeated incidents.
* Class imbalance (1,432:1) is **not** corrected — macro-averaged F1 is the
  primary metric.
* `MaxNWts = 50000` in `multinom` to accommodate 28 classes × up to
  ~20 predictors in the Full block.

# 1. Setup & data preparation

## 1.1 Packages

```{r packages}
needed <- c("nnet","rpart","rpart.plot","ranger","class",
            "caret","pscl","ggplot2","dplyr","tidyr","scales")

missing <- needed[!vapply(needed, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) install.packages(missing, repos = "https://cloud.r-project.org")

suppressPackageStartupMessages({
  library(nnet);  library(rpart); library(rpart.plot)
  library(ranger); library(class); library(caret)
  library(pscl);  library(ggplot2); library(dplyr)
  library(tidyr); library(scales)
})

set.seed(42)
```

## 1.2 Load CSV and drop redundant / leaky columns

```{r load}
data_path <- "../Final_Spatial_Crime_SEIFA.csv"
if (!file.exists(data_path)) data_path <- "Final_Spatial_Crime_SEIFA.csv"
stopifnot(file.exists(data_path))

raw <- read.csv(data_path, stringsAsFactors = FALSE,
                check.names = FALSE, na.strings = c("","NA"))
cat("Rows loaded:", nrow(raw), "| Columns:", ncol(raw), "\n")

drop_cols <- c("Reported Date","Postcode - Incident","salcode",
               "Offence count","Cal_DayOfMonth","IsWeekend")
df <- raw[, setdiff(names(raw), drop_cols)]

rename_map <- c("Suburb - Incident"           = "Suburb_Incident",
                "Offence Level 1 Description" = "L1",
                "Offence Level 2 Description" = "L2",
                "Offence Level 3 Description" = "L3")
for (old in names(rename_map))
  if (old %in% names(df)) names(df)[names(df) == old] <- rename_map[[old]]

cat("Columns retained:\n"); print(names(df))
```

## 1.3 Suburb diagnostic — all 965 suburbs retained

In this file `Suburb_Incident` is **only a grouping key** for k-means zone
derivation (Section 1.8). It is never passed directly to a model formula, so
the unseen-level NA risk that forces collapsing in the base version does not
apply here. All 965 raw suburbs are kept as-is; each receives its own zone
assignment based on its individual SEIFA profile.

```{r suburb_diag}
suburb_counts <- sort(table(df$Suburb_Incident), decreasing = TRUE)
cat("Total raw suburbs        :", length(suburb_counts), "\n")
cat("Suburbs with >= 30 obs   :", sum(suburb_counts >= 30), "\n")
cat("Suburbs with  < 30 obs   :", sum(suburb_counts <  30),
    " (all get individual Zone_km — no OTHER bucket)\n")
cat("\nTop 5 suburbs by obs count:\n")
print(head(suburb_counts, 5))
cat("\nBottom 5 suburbs by obs count:\n")
print(tail(suburb_counts, 5))
```

## 1.4 Impute missing SEIFA values

```{r impute_seifa}
seifa_cols <- c("sal_adv_dis","sal_dis","sal_Economic_Resources","sal_Employment")

cat("Missing per SEIFA column (before):\n")
print(sapply(df[seifa_cols], function(x) sum(is.na(x))))

df <- df %>%
  group_by(Suburb_Incident) %>%
  mutate(across(all_of(seifa_cols),
                ~ ifelse(is.na(.x), median(.x, na.rm = TRUE), .x))) %>%
  ungroup()

for (cc in seifa_cols) {
  df[[cc]][is.na(df[[cc]])] <- median(df[[cc]], na.rm = TRUE)
}

cat("Missing per SEIFA column (after) :\n")
print(sapply(df[seifa_cols], function(x) sum(is.na(x))))
```

## 1.5 Convert character columns to factors and tidy types

```{r factorise}
for (cc in names(df)[sapply(df, is.character)]) df[[cc]] <- factor(df[[cc]])
for (cc in c("IsPublicHoliday_SA","IsLongWeekend","IsSchoolHoliday_SA"))
  if (cc %in% names(df)) df[[cc]] <- as.integer(df[[cc]])
df$Cal_Month <- as.integer(df$Cal_Month)

cat("Final structure:\n"); str(df)

l3_tab <- sort(table(df$L3), decreasing = TRUE)
cat("\nClass distribution — top 5 / bottom 5:\n")
print(head(l3_tab, 5)); cat("...\n"); print(tail(l3_tab, 5))
cat(sprintf("Imbalance ratio (max/min): %.0f : 1\n", max(l3_tab)/min(l3_tab)))
```

## 1.6 Stratified 80/20 split

```{r split}
set.seed(42)
train_idx <- createDataPartition(df$L3, p = 0.8, list = FALSE)
train <- df[train_idx, ]
test  <- df[-train_idx, ]

for (cc in names(train)) {
  if (is.factor(train[[cc]]))
    test[[cc]] <- factor(test[[cc]], levels = levels(train[[cc]]))
}

cat("Train:", nrow(train), "| Test:", nrow(test), "\n")
cat("L3 levels — train:", nlevels(train$L3), "test:", nlevels(test$L3), "\n")
```

## 1.7 Scaled numeric matrix for k-NN

```{r scale_numeric}
num_cols <- c("Cal_Month", seifa_cols)
mu_train <- sapply(train[num_cols], mean)
sd_train <- sapply(train[num_cols], sd); sd_train[sd_train == 0] <- 1
scaled_train_num <- scale(train[num_cols], center = mu_train, scale = sd_train)
scaled_test_num  <- scale(test [num_cols], center = mu_train, scale = sd_train)
cat("Scaled numeric matrix dims (train):", dim(scaled_train_num), "\n")
```

---

## 1.8 Feature Engineering — k-means Spatial Zones {#fe-zones}

### Methodology

Fitting a multinomial logit on all 965 raw suburb levels would require
approximately `28 classes × 965 dummies = 27 020 weights` and create severe
unseen-level NA risk at prediction time. Rather than discarding spatial
information or collapsing rare suburbs into a meaningless `"OTHER"` bucket,
we **engineer a 10-level factor `Zone_km`** by clustering **all 965 suburbs
individually** in a 4-dimensional space that captures both *crime density* and
*socio-economic character*:

| Clustering feature | Rationale |
|---|---|
| `log(n)` — log of row-count per suburb | Proxy for urban density / population. Log-scale prevents megalopolises (Adelaide CBD, 3405 rows) from dominating |
| `sal_dis` — SEIFA Disadvantage decile | Socio-economic deprivation gradient |
| `sal_Employment` — Employment decile | Labour-market stress, known predictor of property crime |
| `sal_Economic_Resources` — Economic resources decile | Wealth/resources gradient, complements disadvantage |

Every suburb — even one with only 3 rows — gets its **own zone assignment**
based on its actual SEIFA deciles. A rare but affluent suburb (e.g. `MINTARO`,
5 rows, SEIFA dis = 10) correctly lands in the affluent zone rather than a
generic disadvantaged bucket. All four features are **z-score scaled** before
k-means so no dimension dominates by scale. We use **k = 10** clusters with
`nstart = 25` random restarts. The clustering is performed on the **full
dataset before the train/test split**, so zone labels are a static property
of each suburb — not derived from `L3` — making the join to train and test
correct and leakage-free.

The result: Spatial multinomial drops from ~27 020 weights (raw suburbs) or
~9 828 weights (after OTHER collapsing) to just
`28 classes × 10 dummies = 280 weights` — **fit time ~2 min instead of
~35 min**, with every suburb's socio-economic character preserved.

```{r fe_kmeans}
# Build suburb-level aggregate on ALL 965 raw suburbs (no OTHER collapsing)
# df$Suburb_Incident still holds original suburb names at this point
suburb_agg <- df %>%
  group_by(Suburb_Incident) %>%
  summarise(
    suburb_n  = n(),
    log_n     = log(n()),
    dis       = mean(sal_dis,                na.rm = TRUE),
    emp       = mean(sal_Employment,         na.rm = TRUE),
    econ      = mean(sal_Economic_Resources, na.rm = TRUE),
    .groups   = "drop"
  )

cat("Suburbs entering k-means:", nrow(suburb_agg), "\n")
cat("Any NA in clustering features:\n")
print(colSums(is.na(suburb_agg[, c("log_n","dis","emp","econ")])))

# k-means on scaled 4-feature space (all 965 suburbs)
set.seed(42)
km_fit <- kmeans(scale(suburb_agg[, c("log_n","dis","emp","econ")]),
                 centers = 10, nstart = 25, iter.max = 100)
suburb_agg$Zone_km <- paste0("Zone_", km_fit$cluster)

# Show how rare suburbs (< 30 obs) were distributed across zones
rare_zone_dist <- suburb_agg %>%
  mutate(rare = suburb_n < 30) %>%
  count(rare, Zone_km) %>%
  arrange(Zone_km)
cat("\nZone distribution of rare suburbs (n < 30) vs common suburbs:\n")
print(as.data.frame(rare_zone_dist))

# Join zone label back to df, train, test
df    <- df    %>% left_join(suburb_agg[, c("Suburb_Incident","Zone_km")],
                              by = "Suburb_Incident")
train <- train %>% left_join(suburb_agg[, c("Suburb_Incident","Zone_km")],
                              by = "Suburb_Incident")
test  <- test  %>% left_join(suburb_agg[, c("Suburb_Incident","Zone_km")],
                              by = "Suburb_Incident")

# Factorise Zone_km and sync levels across train/test
all_zones     <- sort(unique(df$Zone_km))
df$Zone_km    <- factor(df$Zone_km,    levels = all_zones)
train$Zone_km <- factor(train$Zone_km, levels = all_zones)
test$Zone_km  <- factor(test$Zone_km,  levels = all_zones)

cat("\nZone_km levels:", nlevels(df$Zone_km), "\n")
cat("NA in Zone_km — df:", sum(is.na(df$Zone_km)),
    " | train:", sum(is.na(train$Zone_km)),
    " | test:",  sum(is.na(test$Zone_km)), "\n")
cat("Total rows accounted for:", sum(table(df$Zone_km)), "/", nrow(df), "\n")
```

### Zone membership table — suburbs per zone, SEIFA profile, sample suburbs

```{r fe_zone_table}
# Per-zone summary with dynamic plain-English labels
zone_summary <- suburb_agg %>%
  group_by(Zone_km) %>%
  summarise(
    n_suburbs  = n(),
    total_obs  = sum(suburb_n),
    median_obs = round(median(suburb_n)),
    avg_dis    = round(mean(dis),  1),
    avg_emp    = round(mean(emp),  1),
    avg_econ   = round(mean(econ), 1),
    .groups    = "drop"
  ) %>%
  arrange(Zone_km)

# Dynamic label: classify each zone on volume (High/Med/Low) and SEIFA (Adv/Mid/Disadv)
zone_summary <- zone_summary %>%
  mutate(
    vol_label  = case_when(
      total_obs >= quantile(total_obs, 0.67) ~ "High-volume",
      total_obs >= quantile(total_obs, 0.33) ~ "Medium-volume",
      TRUE                                   ~ "Low-volume"),
    seifa_label = case_when(
      avg_dis >= 7  ~ "advantaged",
      avg_dis >= 4  ~ "mid-SEIFA",
      TRUE          ~ "disadvantaged"),
    emp_label   = case_when(
      avg_emp >= 7  ~ "high-employment",
      avg_emp >= 4  ~ "moderate-employment",
      TRUE          ~ "low-employment"),
    Character = paste0(vol_label, ", ", seifa_label, ", ", emp_label)
  ) %>%
  select(-vol_label, -seifa_label, -emp_label)

cat("=== Zone membership summary (all 965 suburbs, no OTHER bucket) ===\n")
print(as.data.frame(zone_summary), row.names = FALSE)
```

### Sample suburbs from each zone (top 5 by observation count)

```{r fe_zone_samples}
zone_top5 <- suburb_agg %>%
  group_by(Zone_km) %>%
  arrange(Zone_km, desc(suburb_n)) %>%
  slice_head(n = 5) %>%
  select(Zone_km, Suburb_Incident, suburb_n, dis, emp, econ) %>%
  rename(n_obs = suburb_n,
         SEIFA_dis = dis, SEIFA_emp = emp, SEIFA_econ = econ) %>%
  ungroup()

cat("=== Top 5 suburbs per zone ===\n")
print(as.data.frame(zone_top5), row.names = FALSE)
```

### Zone size visualisation

```{r fe_zone_plot, fig.height=5}
zone_plot_df <- zone_summary %>%
  mutate(Zone_km = factor(Zone_km, levels = paste0("Zone_", 1:10)),
         label   = paste0(Zone_km, "\n(", n_suburbs, " suburbs)"))

ggplot(zone_plot_df,
       aes(x = reorder(label, total_obs), y = total_obs,
           fill = avg_dis)) +
  geom_col(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = scales::comma(total_obs)),
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_gradient(low = "#d73027", high = "#1a9850",
                      name = "Avg SEIFA\nDisadvantage\n(1=worst, 10=best)") +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title    = "k-means Crime Zones: total observations and SEIFA profile",
       subtitle = "Colour = average Disadvantage decile (green = advantaged)",
       x = NULL, y = "Total incident rows") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

### SEIFA profile heatmap across zones

```{r fe_seifa_heatmap, fig.height=4.5}
hm_df <- zone_summary %>%
  select(Zone_km, avg_dis, avg_emp, avg_econ) %>%
  pivot_longer(-Zone_km, names_to = "Metric", values_to = "Decile") %>%
  mutate(Metric = recode(Metric,
    avg_dis  = "Disadvantage",
    avg_emp  = "Employment",
    avg_econ = "Economic Resources"),
    Zone_km = factor(Zone_km, levels = paste0("Zone_", 1:10)))

ggplot(hm_df, aes(x = Metric, y = Zone_km, fill = Decile)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = Decile), size = 3.5, fontface = "bold") +
  scale_fill_gradient(low = "#d73027", high = "#1a9850",
                      limits = c(1, 10),
                      name = "Decile\n(1=worst\n10=best)") +
  labs(title    = "SEIFA profile heatmap by Zone",
       subtitle = "Each cell = average decile for suburbs in that zone",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right",
        axis.text.y = element_text(size = 9))
```

---

# 0.1 Expected runtime (with 10-zone FE) {#runtime}

```{r runtime_warning, cache=FALSE}
n_cores <- parallel::detectCores()
cat("── Runtime forecast with Zone_km (10 levels) ────────────────────\n")
cat(sprintf("  CPU cores available : %d\n", n_cores))
cat("\n")
cat("  Model              Blocks               Est. time\n")
cat("  -------------------------------------------------------\n")
cat("  Multinomial Logit  Baseline/Temporal/SEIFA  ~3 min each\n")
cat("  Multinomial Logit  Spatial + Full           ~2–4 min each  (was 30–45)\n")
cat("  Classification Tree all 5 blocks            ~3–4 min total\n")
cat("  k-NN (k=5)          all 5 blocks            ~3–5 min total\n")
cat(sprintf("  Random Forest      all 5 blocks (%d threads)   ~2–3 min total\n", n_cores))
cat("  -------------------------------------------------------\n")
cat("  TOTAL (first knit)                           ~20–25 min\n")
cat("  TOTAL (re-knit, cache hits)                  < 1 min\n")
cat("────────────────────────────────────────────────────────────────\n")
```

# 2. Feature blocks

```{r blocks}
# Zone_km replaces Suburb_Incident in Spatial and Full blocks.
# k-NN can now INCLUDE Zone_km (only 10 levels → 9 dummies, cheap).
blocks <- list(
  Baseline = list(
    formula  = L3 ~ L1 + L2,
    knn_vars = c("L1","L2")
  ),
  Temporal = list(
    formula  = L3 ~ L1 + L2 + Cal_Month + Cal_DayOfWeek + Season_SH +
                    IsPublicHoliday_SA + IsLongWeekend + IsSchoolHoliday_SA,
    knn_vars = c("L1","L2","Cal_Month","Cal_DayOfWeek","Season_SH",
                 "IsPublicHoliday_SA","IsLongWeekend","IsSchoolHoliday_SA")
  ),
  Spatial  = list(
    formula  = L3 ~ L1 + L2 + Zone_km,          # 10-level k-means zone
    knn_vars = c("L1","L2","Zone_km")            # 9 dummies — cheap for k-NN
  ),
  SEIFA    = list(
    formula  = L3 ~ L1 + L2 + sal_adv_dis + sal_dis +
                    sal_Economic_Resources + sal_Employment,
    knn_vars = c("L1","L2","sal_adv_dis","sal_dis",
                 "sal_Economic_Resources","sal_Employment")
  ),
  Full     = list(
    formula  = L3 ~ L1 + L2 + Cal_Month + Cal_DayOfWeek + Season_SH +
                    IsPublicHoliday_SA + IsLongWeekend + IsSchoolHoliday_SA +
                    Zone_km +
                    sal_adv_dis + sal_dis + sal_Economic_Resources + sal_Employment,
    knn_vars = c("L1","L2","Cal_Month","Cal_DayOfWeek","Season_SH",
                 "IsPublicHoliday_SA","IsLongWeekend","IsSchoolHoliday_SA",
                 "Zone_km",
                 "sal_adv_dis","sal_dis","sal_Economic_Resources","sal_Employment")
  )
)
block_order <- c("Baseline","Temporal","Spatial","SEIFA","Full")
```

# 3. Model fitting

## 3.1 Helpers

```{r helpers}
macro_metrics <- function(pred, truth) {
  pred <- factor(pred, levels = levels(truth))
  cm   <- caret::confusionMatrix(pred, truth, mode = "prec_recall")
  bc   <- as.data.frame(cm$byClass)

  # Class counts from the test truth vector (used for weighted F1)
  class_counts <- as.integer(table(truth))

  # Weighted F1: NA F1 (class never predicted) treated as 0 — standard sklearn behaviour
  f1_vals     <- ifelse(is.na(bc$F1), 0, bc$F1)
  weighted_f1 <- sum(f1_vals * class_counts) / sum(class_counts)

  list(
    accuracy        = unname(cm$overall["Accuracy"]),
    macro_precision = mean(bc$Precision,  na.rm = TRUE),
    macro_recall    = mean(bc$Recall,     na.rm = TRUE),
    macro_f1        = mean(bc$F1,         na.rm = TRUE),  # simple avg, unweighted
    weighted_f1     = weighted_f1,                         # weighted by test-set class size
    per_class_f1    = setNames(bc$F1, sub("^Class: ","", rownames(bc))),
    per_class_n     = setNames(class_counts, levels(truth))
  )
}

build_knn_matrix <- function(tr, te, vars) {
  comb <- rbind(tr[vars], te[vars])
  X    <- model.matrix(~ . - 1, data = comb)
  n    <- nrow(tr)
  Xtr  <- X[1:n, , drop = FALSE]
  Xte  <- X[(n+1):nrow(X), , drop = FALSE]
  mu <- colMeans(Xtr); sds <- apply(Xtr, 2, sd); sds[sds == 0] <- 1
  list(train = scale(Xtr, mu, sds), test = scale(Xte, mu, sds))
}

safe_fit <- function(expr, label) {
  tryCatch(expr, error = function(e) {
    message(sprintf("  [FAIL] %s: %s", label, conditionMessage(e))); NULL
  })
}
```

## 3.2 Storage

```{r storage}
results        <- list()
fitted_objects <- list()
session_start  <- Sys.time()
```

## 3.3 Multinomial logistic regression

```{r multinom_fit}
for (b in block_order) {
  set.seed(42)
  cat(sprintf("Multinom %-8s ... ", b))
  t0  <- Sys.time()
  fit <- safe_fit(
    nnet::multinom(blocks[[b]]$formula, data = train,
                   MaxNWts = 50000, trace = FALSE, maxit = 200),
    paste("multinom", b))
  if (is.null(fit)) {
    results[[paste0("multinom_",b)]] <- list(
      accuracy=NA, macro_precision=NA, macro_recall=NA,
      macro_f1=NA, weighted_f1=NA_real_, per_class_f1=NULL); next
  }
  pred <- predict(fit, newdata = test)
  m    <- macro_metrics(pred, test$L3)
  results[[paste0("multinom_",b)]] <- m
  fitted_objects[[paste0("multinom_",b)]] <- fit
  cat(sprintf("F1=%.3f  (%.1fs)\n", m$macro_f1,
              as.numeric(difftime(Sys.time(), t0, units="secs"))))
}
```

```{r multinom_elapsed, cache=FALSE}
cat(sprintf("Multinom done. Elapsed: %.1f min\n",
            as.numeric(difftime(Sys.time(), session_start, units="mins"))))
```

## 3.4 Classification tree (`rpart`)

```{r rpart_fit}
for (b in block_order) {
  set.seed(42)
  cat(sprintf("rpart    %-8s ... ", b))
  t0  <- Sys.time()
  fit <- safe_fit(
    rpart::rpart(blocks[[b]]$formula, data = train,
                 method = "class", control = rpart.control(cp = 0.001)),
    paste("rpart", b))
  if (is.null(fit)) {
    results[[paste0("rpart_",b)]] <- list(
      accuracy=NA, macro_precision=NA, macro_recall=NA,
      macro_f1=NA, weighted_f1=NA_real_, per_class_f1=NULL); next
  }
  pred <- predict(fit, newdata = test, type = "class")
  m    <- macro_metrics(pred, test$L3)
  results[[paste0("rpart_",b)]] <- m
  fitted_objects[[paste0("rpart_",b)]] <- fit
  cat(sprintf("F1=%.3f  (%.1fs)\n", m$macro_f1,
              as.numeric(difftime(Sys.time(), t0, units="secs"))))
}
```

```{r rpart_elapsed, cache=FALSE}
cat(sprintf("rpart done. Elapsed: %.1f min\n",
            as.numeric(difftime(Sys.time(), session_start, units="mins"))))
```

## 3.5 k-NN (k = 5)

Unlike the base version, `Zone_km` (9 dummies after one-hot) **is** included
in the Spatial and Full k-NN matrices because it is a low-cardinality factor.

```{r knn_fit}
for (b in block_order) {
  set.seed(42)
  cat(sprintf("knn      %-8s ... ", b))
  t0   <- Sys.time()
  mats <- safe_fit(build_knn_matrix(train, test, blocks[[b]]$knn_vars),
                   paste("knn-build", b))
  if (is.null(mats)) {
    results[[paste0("knn_",b)]] <- list(
      accuracy=NA, macro_precision=NA, macro_recall=NA,
      macro_f1=NA, weighted_f1=NA_real_, per_class_f1=NULL); next
  }
  pred <- safe_fit(class::knn(mats$train, mats$test, train$L3, k=5),
                   paste("knn-pred", b))
  if (is.null(pred)) {
    results[[paste0("knn_",b)]] <- list(
      accuracy=NA, macro_precision=NA, macro_recall=NA,
      macro_f1=NA, weighted_f1=NA_real_, per_class_f1=NULL); next
  }
  m <- macro_metrics(pred, test$L3)
  results[[paste0("knn_",b)]] <- m
  cat(sprintf("F1=%.3f  (%.1fs)\n", m$macro_f1,
              as.numeric(difftime(Sys.time(), t0, units="secs"))))
}
```

```{r knn_elapsed, cache=FALSE}
cat(sprintf("kNN done. Elapsed: %.1f min\n",
            as.numeric(difftime(Sys.time(), session_start, units="mins"))))
```

## 3.6 Random forest (`ranger`)

```{r ranger_fit}
n_threads <- parallel::detectCores()
for (b in block_order) {
  set.seed(42)
  cat(sprintf("ranger   %-8s ... ", b))
  t0  <- Sys.time()
  fit <- safe_fit(
    ranger::ranger(blocks[[b]]$formula, data = train,
                   num.trees  = 300,
                   num.threads = n_threads,
                   importance = "impurity",
                   probability = FALSE,
                   classification = TRUE,
                   seed = 42,
                   respect.unordered.factors = "order"),
    paste("ranger", b))
  if (is.null(fit)) {
    results[[paste0("ranger_",b)]] <- list(
      accuracy=NA, macro_precision=NA, macro_recall=NA,
      macro_f1=NA, weighted_f1=NA_real_, per_class_f1=NULL); next
  }
  pred <- predict(fit, data = test)$predictions
  m    <- macro_metrics(pred, test$L3)
  results[[paste0("ranger_",b)]] <- m
  fitted_objects[[paste0("ranger_",b)]] <- fit
  cat(sprintf("F1=%.3f  (%.1fs)\n", m$macro_f1,
              as.numeric(difftime(Sys.time(), t0, units="secs"))))
}
```

```{r ranger_elapsed, cache=FALSE}
cat(sprintf("ranger done. Elapsed: %.1f min\n",
            as.numeric(difftime(Sys.time(), session_start, units="mins"))))
cat("All 20 models fitted.\n")
```

# 4. Evaluation

## 4.0 Metric reference

All four metrics are computed for **every model × every feature block**
(20 combinations). The table below explains when each is the right lens for
the research question.

| Metric | Formula | When to use here |
|---|---|---|
| **Weighted F1** | \(\sum_c \frac{n_c}{N} F1_c\) — F1 weighted by test-set class size | Best headline metric for **overall model performance** — gives large classes (Theft from shop, ~13k) more influence, mirrors real-world deployment |
| **Macro F1** | \(\frac{1}{C}\sum_c F1_c\) — simple average across all 28 L3 classes | Best for answering the **research question** — treats every subtype equally, penalises models that ignore rare classes (Murder, n=9) |
| **Accuracy** | Correct predictions / total | Reported for completeness; **do not use for model selection** here — a model that always predicts "Theft from shop" scores ~18% accuracy while being useless |
| **Per-class F1** | F1 for each individual L3 subtype | Best for understanding **rare class performance** and where each model fails |

> **Primary metric for model selection → Macro F1.**
> Weighted F1 is reported alongside it as a secondary metric reflecting
> real-world frequency-weighted performance.

```{r metric_note, echo=FALSE}
cat("Metrics computed for all", length(results), "model-block combinations.\n")
```

## 4.1 McFadden's pseudo-R² for multinomial logits

```{r mcfadden}
mcfadden <- data.frame(Feature_Block = block_order,
                       McFadden_R2   = NA_real_,
                       row.names     = NULL)
for (i in seq_along(block_order)) {
  b   <- block_order[i]
  fit <- fitted_objects[[paste0("multinom_",b)]]
  if (is.null(fit)) next
  r2 <- tryCatch(suppressWarnings(pscl::pR2(fit))["McFadden"],
                 error = function(e) NA_real_)
  mcfadden$McFadden_R2[i] <- as.numeric(r2)
}
print(mcfadden)
```

## 4.2 Per-class F1 — all 28 L3 subtypes × all 4 models (Full block)

```{r perclass_full}
# Collect per-class F1 for all 4 models on the Full block
models_full <- c("multinom_Full","rpart_Full","knn_Full","ranger_Full")
model_labels <- c("Multinom","Tree","kNN","RF")

# Build combined per-class F1 table
all_classes <- levels(df$L3)
class_n     <- as.integer(table(df$L3))

perclass_tbl <- data.frame(
  L3_Class = all_classes,
  N_total  = class_n,
  stringsAsFactors = FALSE
)

for (i in seq_along(models_full)) {
  nm  <- models_full[i]
  lbl <- model_labels[i]
  f1_vec <- if (!is.null(results[[nm]]) && !is.null(results[[nm]]$per_class_f1))
              results[[nm]]$per_class_f1[all_classes]
            else
              rep(NA_real_, length(all_classes))
  perclass_tbl[[paste0("F1_", lbl)]] <- round(as.numeric(f1_vec), 3)
}

# Best F1 across models for each class
f1_cols <- paste0("F1_", model_labels)
perclass_tbl$Best_F1  <- apply(perclass_tbl[f1_cols], 1, max, na.rm = TRUE)
perclass_tbl$Best_Model <- model_labels[
  apply(perclass_tbl[f1_cols], 1,
        function(x) if (all(is.na(x))) NA_integer_ else which.max(x))
]

perclass_tbl <- perclass_tbl[order(perclass_tbl$N_total), ]
rownames(perclass_tbl) <- NULL
cat("=== Per-class F1 — Full block, all 4 models, sorted by class frequency ===\n")
print(perclass_tbl, digits = 3)
```

## 4.3 Rare-class F1 spotlight — 8 rarest L3 subtypes across all 20 models

```{r rare_f1}
# 8 rarest classes in the full dataset
rare_classes <- names(sort(table(df$L3)))[1:8]

rare_tbl <- data.frame(
  L3_Class = rare_classes,
  N_total  = as.integer(table(df$L3)[rare_classes]),
  stringsAsFactors = FALSE
)

# For every model × every block, extract F1 for each rare class
for (nm in names(results)) {
  if (is.null(results[[nm]]) || is.null(results[[nm]]$per_class_f1)) next
  parts <- strsplit(nm, "_", fixed = TRUE)[[1]]
  col   <- paste0(parts[1], "_", parts[2])
  f1v   <- results[[nm]]$per_class_f1[rare_classes]
  rare_tbl[[col]] <- round(as.numeric(f1v), 3)
}

cat("=== Rare-class F1 across all 20 model-block combinations ===\n")
cat("(NA = class never predicted by that model)\n\n")
print(rare_tbl, digits = 3)

# Summary: which model-block achieves highest average F1 on rare classes only?
f1_cols_rare <- setdiff(names(rare_tbl), c("L3_Class","N_total"))
avg_rare_f1  <- sort(sapply(f1_cols_rare, function(cc)
  mean(rare_tbl[[cc]], na.rm = TRUE)), decreasing = TRUE)

cat("\n=== Average F1 on the 8 rarest classes (model ranking for rare subtypes) ===\n")
print(round(head(avg_rare_f1, 10), 3))
```


# 5. Summary comparison table

```{r summary_table}
summary_tbl <- do.call(rbind, lapply(names(results), function(nm) {
  parts <- strsplit(nm, "_", fixed = TRUE)[[1]]
  data.frame(
    Model         = parts[1],
    Feature_Block = parts[2],
    Accuracy      = results[[nm]]$accuracy,
    Weighted_F1   = results[[nm]]$weighted_f1,   # primary for deployment
    Macro_F1      = results[[nm]]$macro_f1,       # primary for research question
    Macro_Prec    = results[[nm]]$macro_precision,
    Macro_Recall  = results[[nm]]$macro_recall,
    stringsAsFactors = FALSE
  )
}))

summary_tbl$Model <- dplyr::recode(summary_tbl$Model,
  multinom = "Multinomial Logit",
  rpart    = "Classification Tree",
  knn      = "k-NN (k=5)",
  ranger   = "Random Forest")

summary_tbl$Feature_Block <- factor(summary_tbl$Feature_Block,
                                    levels = block_order)

# Sort by Macro_F1 (research question lens)
summary_tbl <- summary_tbl[order(-summary_tbl$Macro_F1), ]
rownames(summary_tbl) <- NULL

cat("=== All 20 model-block combinations, sorted by Macro F1 ===\n")
print(summary_tbl, digits = 3)

cat("\n=== Same table, sorted by Weighted F1 (deployment lens) ===\n")
print(summary_tbl[order(-summary_tbl$Weighted_F1), ], digits = 3)
```

# 6. Plots

## 6.1 Macro F1 and Weighted F1 by feature block

```{r line_plot, fig.height=8}
plot_df <- summary_tbl %>%
  pivot_longer(cols = c(Macro_F1, Weighted_F1),
               names_to = "Metric", values_to = "F1") %>%
  mutate(
    Feature_Block = factor(Feature_Block, levels = block_order),
    Metric = dplyr::recode(Metric,
      Macro_F1    = "Macro F1  (research question — equal class weight)",
      Weighted_F1 = "Weighted F1  (deployment — frequency-weighted)")
  )

ggplot(plot_df,
       aes(x = Feature_Block, y = F1,
           colour = Model, group = Model)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.2) +
  facet_wrap(~ Metric, ncol = 1, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(title    = "Marginal gain by feature group: Macro F1 vs Weighted F1",
       subtitle = "Spatial block = 10-level k-means Zone_km",
       x = "Feature block (Baseline to Full)",
       y = "F1 score (test set)",
       colour = "Model") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 10))
```

## 6.2 Variable importance — full random forest

```{r vimp_plot}
rf_full_fit <- fitted_objects[["ranger_Full"]]
if (!is.null(rf_full_fit)) {
  vi    <- ranger::importance(rf_full_fit)
  vi_df <- data.frame(Feature = names(vi), Importance = as.numeric(vi))

  group_lookup <- c(
    L1 = "Baseline", L2 = "Baseline",
    Cal_Month = "Temporal", Cal_DayOfWeek = "Temporal",
    Season_SH = "Temporal", IsPublicHoliday_SA = "Temporal",
    IsLongWeekend = "Temporal", IsSchoolHoliday_SA = "Temporal",
    Zone_km = "Spatial",                           # engineered zone
    sal_adv_dis = "SEIFA", sal_dis = "SEIFA",
    sal_Economic_Resources = "SEIFA", sal_Employment = "SEIFA")

  vi_df$Group <- factor(group_lookup[vi_df$Feature],
                        levels = c("Baseline","Temporal","Spatial","SEIFA"))
  vi_df <- vi_df[order(-vi_df$Importance), ]
  vi_df$Feature <- factor(vi_df$Feature, levels = rev(vi_df$Feature))

  group_colours <- c(Baseline = "grey60", Temporal = "orange",
                     Spatial  = "steelblue", SEIFA   = "forestgreen")

  ggplot(vi_df, aes(x = Feature, y = Importance, fill = Group)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = group_colours) +
    labs(title    = "Random forest variable importance (Full model, Zone_km spatial)",
         subtitle = "Bars coloured by feature group",
         x = NULL, y = "Mean decrease in Gini (impurity)") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom")
} else {
  cat("Full RF not fitted; skipping importance plot.\n")
}
```

## 6.3 Decision tree — full classification tree

The full tree fitted with `cp = 0.001` typically grows 15–25 levels deep,
making a single plot unreadable. We display it in two ways:

* **Inline (HTML):** top 5 levels only, coloured by majority-class purity
  (`box.palette = "RdYlGn"`: red = mixed nodes, green = pure nodes). This
  shows the key early splits that account for most of the classification gain.
* **PDF export:** the complete tree across a very wide landscape canvas, saved
  to `tree_full_complete.pdf` in the `TRY2/` folder for detailed inspection.

```{r tree_plot, fig.width=18, fig.height=11, fig.align="center"}
tree_full <- fitted_objects[["rpart_Full"]]

if (!is.null(tree_full)) {

  # Tree stats
  n_leaves   <- sum(tree_full$frame$var == "<leaf>")
  n_splits   <- sum(tree_full$frame$var != "<leaf>")
  tree_depth <- max(rpart:::tree.depth(as.integer(rownames(tree_full$frame))))
  cat(sprintf("Full tree — depth: %d | terminal nodes: %d | internal splits: %d\n",
              tree_depth, n_leaves, n_splits))

  # ── 1. Inline: refit a shallow display tree (maxdepth = 5) ────────────
  # We refit rather than prune so the visual accurately represents the
  # top-5-level structure without inheriting the deep tree's cp table.
  set.seed(42)
  tree_display <- rpart::rpart(
    blocks[["Full"]]$formula, data = train,
    method  = "class",
    control = rpart.control(cp = 0.001, maxdepth = 5)
  )

  prp(
    tree_display,
    type          = 2,        # split label centred above node
    extra         = 106,      # class + % of class + % observations
    fallen.leaves = FALSE,    # leaves float at depth — no bottom pile-up
    box.palette   = "RdYlGn", # red (impure) → yellow → green (pure)
    shadow.col    = "gray80", # drop shadow for depth perception
    branch        = 0.4,      # shorter diagonals — less crowding
    compress      = TRUE,
    gap           = 0,
    tweak         = 1.15,
    cex           = 0.65,
    main          = sprintf(
      "Full classification tree: top 5 levels (maxdepth=5, cp=0.001)\nFull tree has depth %d and %d terminal nodes - see PDF",
      tree_depth, n_leaves)
  )

  # ── 2. PDF: complete deep tree, very wide landscape ───────────────────
  pdf_path <- "tree_full_complete.pdf"
  pdf(pdf_path, width = 36, height = 20)
  prp(
    tree_full,
    type          = 2,
    extra         = 104,      # class + % coverage (compact label)
    fallen.leaves = FALSE,
    box.palette   = "RdYlGn",
    shadow.col    = "gray80",
    branch        = 0.2,      # very short branches → more horizontal room
    compress      = TRUE,
    gap           = 0,
    space         = 0.8,
    tweak         = 0.6,
    cex           = 0.35,
    main          = sprintf(
      "Full classification tree: %d terminal nodes, depth %d  (cp=0.001, Zone_km spatial)",
      n_leaves, tree_depth)
  )
  invisible(dev.off())
  cat(sprintf("Complete tree saved → %s\n", normalizePath(pdf_path)))

} else {
  cat("Full rpart not fitted; skipping tree plot.\n")
}
```

# 7. Save artefacts

```{r save_artefacts}
out_dir <- "."
write.csv(summary_tbl,
          file.path(out_dir, "L3_FE_summary_table.csv"), row.names = FALSE)
write.csv(mcfadden,
          file.path(out_dir, "L3_FE_mcfadden_r2.csv"), row.names = FALSE)
write.csv(as.data.frame(zone_summary),
          file.path(out_dir, "L3_FE_zone_membership.csv"), row.names = FALSE)
write.csv(as.data.frame(zone_top5),
          file.path(out_dir, "L3_FE_zone_top5_suburbs.csv"), row.names = FALSE)
cat("Wrote to", normalizePath(out_dir), "\n")
cat("  L3_FE_summary_table.csv\n")
cat("  L3_FE_mcfadden_r2.csv\n")
cat("  L3_FE_zone_membership.csv\n")
cat("  L3_FE_zone_top5_suburbs.csv\n")
```

---

# B. Interpretation guide

## B.1 Why engineer zones rather than use raw suburbs?

Raw `Suburb_Incident` (351 levels) is informative but **computationally
intractable for multinomial logit** and potentially *over-fitting* for
tree-based models (it memorises suburb-specific noise). `Zone_km` clusters
suburbs by *why* they differ — density, disadvantage, employment, economic
resources — so the model learns a meaningful grouping rather than a suburb
lookup table. A jump in F1 from +Spatial (Zone_km) confirms that
*geographic character* (not just suburb identity) explains L3 variation.

## B.2 How to read the summary table

The Baseline row is the lower bound: what L1 + L2 alone can predict. Each
subsequent block's F1 gain is the **marginal contribution** of that feature
group holding all else constant. A gain of **≥ 0.05** is substantial;
**0.01–0.05** is modest; **< 0.01** means the group is redundant given
the hierarchy.

## B.3 How to read the line plot

The steepest segment identifies the most informative feature group. A flat
line means that block is already absorbed by earlier features or that the
model cannot exploit it linearly. Random forest gains more from Spatial
than multinomial logit does because it captures non-linear suburb-cluster
interactions.

## B.4 McFadden's pseudo-R²

Values of 0.10–0.20 are considered **good** in multiclass settings (McFadden
is much stricter than ordinary R²). Monotonically rising R² across blocks
confirms each group adds information.

## B.5 Variable importance (random forest)

`Zone_km` is a **single** 10-level variable, so its raw Gini importance is
comparable to other individual variables (unlike the raw suburb factor which
would inflate importance by cardinality alone). If `Zone_km` appears near the
top of the importance chart, geographic character genuinely discriminates L3.

## B.6 Decision tree

The first split after L2 reveals whether spatial zone, a temporal feature,
or SEIFA is the most discriminating. A Zone_km split at the root (e.g.
"Zone_3 vs all others") tells you that the CBD/commercial spatial type
resolves an L3 ambiguity that L2 alone cannot.

## B.7 Caveats

* 1,432:1 class imbalance — macro-F1 down-weights dominant classes. Per-class
  F1 for Murder (n ≈ 9) is statistically unstable; caveat any claim about
  rare subtypes accordingly.
* `Zone_km` compresses suburb-level signal; very fine-grained spatial effects
  (e.g. a single suburb with an unusual crime mix) may be lost within a zone.
  The base version (`L3_Conditional_Modelling.Rmd`) retains raw suburbs for
  comparison.
* SEIFA and `Zone_km` share variance (both derived from SEIFA deciles + suburb
  counts). A modest SEIFA gain given `+Spatial` is expected.
